Evaluation of 3-point correlation functions from structural images on CPU and GPU architectures: accounting for anisotropy effects

### 2024-08-29 paper.tex
last changes before revision

\author{Vasily~Postnicov\textsuperscript{1}}
\author{Marina~V.~Karsanina\textsuperscript{1}}
\author{Aleksey~Khlyupin\textsuperscript{1,2}}
\author{Kirill~M.~Gerke\textsuperscript{1}}
\email{kg@ifz.ru}

\affiliation{\textsuperscript{1}Schmidt Institute of Physics of the Earth of
  Russian Academy of Sciences, Moscow, 107031, Russia}
\affiliation{\textsuperscript{2}Moscow Institute of Physics and Technology,
  Dolgoprudny, 141701, Russia}

@@

\begin{abstract}
Structures, or spatial arrangements of matter and energy, including some fields
(e.g., velocity or pressure) are ubiquitous in research applications and
frequently require description for subsequent analysis, or stochastic
reconstruction from limited data.  The classical descriptors are 2-point
correlation functions (CFs), but the computation of 3-point statistics is known
to be advantageous in some cases as they can probe non-Gaussian signatures, not
captured by their 2-point counterparts. Moreover, $n$-point CFs with $n \ge 3$
are believed to possess larger information content and provide more information
about studied structures. In this paper, we have developed algorithms and code
to compute $S_3, C_3, F_{sss}, F_{ssv}$ and $F_{svv}$ with a right-angle
triangle pattern. This pattern is much easier to vectorize and parallelize
compared to previously used random triangle or spider-web sampling
patterns. Moreover, it allows to easily implement computations of directional
3-point CFs. The execution times for the same number of samples are orders of
magnitude lower than in existing published counterparts. We showed that the volume
of data produced get unwieldy very easily, especially if computations are
performed in frequency domain. For these reasons until information content of
different sets of correlation functions with different "n-pointness" is not
known, advantages of CFs with $n>3$ are not clear. Nonetheless, developed
algorithms and code are universal enough to be easily extendable to any $n$
(with increasing computational and RAM burden). All results are available as
part of open source package \code{CorrelationFunctions.jl} [Postnicov et al.,
  2024, Comp.Phys.Commun., 299, 109134]. As described in this paper, 3-point CFs
computations can be immediately applied in a great number of research
applications, for example: 1) flow and transport velocity fields analysis or any
data with non-Gaussian signatures, 2) deep learning for structural and physical
properties, 3) structure taxonomy and categorization.  In all these and numerous
other potential cases the ability to compute directional 3-point functions may
be crucial. Notably, the organization of the code functions allows computation
of cross-correlation, i.e., one can compute 3-point CFs for multiphase images
(while binary structures were used in this paper for simplicity of
explanations).
\end{abstract}

@@

\section{Introduction}
\label{sec:intro}
Structures, or spatial arrangements of matter and energy, including some fields
(e.g., velocity or pressure) are ubiquitous in research applications and
frequently require description for subsequent analysis. Examples of such objects
for studies include galaxy formations \cite{springel2006}, immiscible multiphase
fluid flow patterns \cite{hopkins2015new,balashov2021}, rock and soil samples
\cite{rozenbaum2014,karsanina2015,ledesma2018,chen2020super,prokhorov2021digital,vogel2010},
food specimens \cite{derossi2019,nagdalian2021}, biological tissue
\cite{park2020}, vortexes during flow \cite{gorbunova2016precessing} and
interstellar turbulence \cite{portillo2018developing}. The scale of interest may
span orders of magnitude from angstroms and nm for materials
\cite{garum2020,gerke2021,khlyupin2023molecular} to millions of light-years for
star clusters \cite{takada2003three,hopkins2013stars}. The studied structure can
be non-static -- it will change with time, and this dynamics is usually also of
great interest \cite{jiao2013,fomin2023soil}. Thus, description of the structure
is of utmost importance for numerous applications:
\begin{enumerate}
  \item Elucidation of structure--property relationships in the form of
    theoretical expression \cite{Torquato_book,Sahimi_book}, e.g., rigorous
    bounds and such;
  \item Prediction of material physical properties using machine (deep) learning
    \cite{obayashi2018persistence,kamrava2020linking,roding2020predicting};
  \item Description of complex structures and fields for taxonomy, comparison
    and analysis
    \cite{takada2003three,hopkins2013stars,shivashankar2015felix,saadatfar2017pore,
      portillo2018developing,cheng2022data,KarsaninaEJSS,PNM_Morse,khlyupin2023molecular},
    which may include temporal dynamics or surface evolution
    \cite{jiao2013,PhysRevE.92.023301,prokhorov2022,chen2022,fomin2023soil};
  \item Morphological representativity
    \cite{capek2009,rozenbaum2014,gerke2019tensor} and spatial stationarity
    analysis \cite{REVpaper,LavrukhinPRE};
  \item Stochastic reconstructions from limited amount of input data
    \cite{Adler_recon,Y-T,tahmasebiPRL,Euras2012,EPL2,karsaninaPRL,rozanski2023}
    in the form of limited dimension structures or correlation functions;
  \item Compress structural information for storage, analysis, training (feature
    extraction), and retrieval (by means of stochastic reconstructions)
    \cite{jiao2007,SciRep1,Havelka,KarsaninaEJSS};
  \item Multiscale image fusion that is usually necessary due to trade-off
    between field-of-view and imaging resolution
    \cite{SciRep1,Geoderma2018,chen2020super,yan2023multiscale,karimpouli2022,yan2024modeling}.
\end{enumerate}

The number of possible descriptors to quantify spatial structure developed
within different disciplines is stunning, for example: phase ratio or porosity
\cite{anovitz2015characterization}, surface area, radial distribution functions
\cite{zimm1948scattering,becker2010radial}, Minkowski functionals
\cite{vogel2010} and tensors \cite{schroder2011minkowski}, correlation functions
\cite{Torquato_book}, persistence diagrams
\cite{shivashankar2015felix,saadatfar2017pore,obayashi2018persistence}. The
evolution of such metrics was mainly induced by the ease of measurement and
interpretation. In fact, we need a single universal descriptor defined by only
one characteristic -- information content. The latter is an ability of the
metric to describe the structure at hand with 100\% meaning the ability to
fully recover the structure back from this metric. Thus, universal descriptor
has to possess following key properties: (1) high information content, (2) ease
of assess, (3) high compression ratio (i.e., descriptor is much "smaller" than
the structure itself).

@@

is based on so-called correlation functions (CFs). These functions define a

@@

    for example, to add functions into a single one \cite{moctezuma2002};

@@

    of higher order $S_n$ are known and were found to be close to direct

@@

\subsection{Open-source implementation}
The described algorithms are implemented in our \code{CorrelationFunctions.jl}
package written in Julia and have the same interface:
\begin{verbatim}
corrfn(input, phase, ps1, ps2)
\end{verbatim}
where \code{corrfn} stands for \code{s3} ($S_3$), \code{c3} ($C_3$),
\code{surf3} ($F_{sss}$) etc., \code{input} is the input array, \code{phase} is
a phase of interest and \code{ps1} and \code{ps2} are arrays of points in which
correlation function is calculated. Firstly, \code{corrfn} does all necessary
preparations for the input array (e.g. edge extraction or connected component
labeling) and then calculates the correlation function for each pair of
arguments from \code{ps1} and \code{ps2}.

Our implementation provides a predefined right triangle pattern where arguments
of a correlation function $x_1$ and $x_2$ are fixed to be parallel to one of the
axes $X$, $Y$ or $Z$ with restriction $x_1 \perp x_2$ (\cref{fig:pattern})
\cite{dimitrakopoulos2010high}. The output has dimensions $D_1 \times D_2$ where
$D_1$ and $D_2$ are dimensions of the input array along selected axes.
\highlight{I CANNOT UNDESTAND THIS SENTENCE: Note that shift operators with such
  $x_1$ and $x_2$ represent a 3-point function computed within a 3D domain, but
  along a predefined direction (see below).}
\begin{figure}[ht]
  \centering
  \includegraphics[width=0.5\linewidth]{images/pattern.png}
  \caption[]{The pattern used for calculation of $S_3$ function in our
    implementation.}
  \label{fig:pattern}
\end{figure}

If it is desired, a user can use a custom pattern more appropriate to the
situation.

\subsection{Analysis of anisotropic media}
When an analyzed sample is anisotropic, i.e., it is stretched or oriented along
some specific direction, it may be beneficial to align one of the cathetus of
sampling triangle \cref{fig:pattern} with that direction. In this case the
anisotropy present in the sample will be properly captured by computed 3-point
correlation function.

To detect the direction of anisotropy we use an $N \times N$ (where $N$ is
dimensionality of the sample) covariance matrix
$\Sigma = \left\{ \sigma_{ij} \right\}$ for coordinates of points in the phase of
interest:
\begin{equation}
  \sigma_{ij} = \sum_{x \in \left\{
    \begin{array}{l}
      \text{Coordinates of points} \\
      \text{in the phase of interset}
    \end{array}
    \right\}} (x_i - \overline{x_i})(x_j - \overline{x_j})
\end{equation}
and decompose $\Sigma$ as $T \Lambda T^{-1}$ where $T$ is a unitary matrix and
$\Lambda$ is diagonal. $T$ can always be made so that it defines a rotation (by
multiplication of one of the columns by $-1$ if needed, so that
$\det T = 1$). Application of this rotation to the sample (or its inverse to the
pattern) makes the direction of anisotropy coaligned with the pattern, so
anisotropy can be better represented in the correlation function. For the sake
of visual clarity in this study we choose to rotate the image instead of the
pattern (\cref{fig:aniso}). This has such consequences as reduced accuracy of
computed functions due to necessity to interpolate the image during
rotation. Although we show in \cref{sec:rotqua} that even nearest-neighbor
interpolation results only in insignificant loss of quality. Rotation of the
pattern has same drawbacks related to interpolation of the sample which are most
noticeable at small correlation lengths. The both methods (rotation of the
sample or the pattern) produce nearly equivalent result if the sample is
homogeneous and an average pore size is much less than the sample's size (see
\cref{sec:rotrot})
\begin{figure*}[!htp]
  \centering
  \begin{tabular}{ccc}
    \includegraphics[width=0.3\linewidth, frame]{images/aniso.png} &
    \includegraphics[width=0.3\linewidth, frame]{images/aniso-rot.png} & \\
    \includegraphics[width=0.3\linewidth]{images/aniso-s3.png} &
    \includegraphics[width=0.3\linewidth]{images/aniso-rot-s3.png} &
    \includegraphics[width=0.3\linewidth]{images/aniso-avg-s3.png} \\
  \end{tabular}
  \caption{Behaviour of our algorithm for calculation of the $S_3$ function
    when dealing with highly anisotropic media. The pattern \cref{fig:pattern}
    which we use in our algorithm is most sensitive to anisotropy which is
    directed along one of the axes. \\
    From top down, left to right: an original image of an anisotropic pore; a
    rotated image of the pore, so it is elongated along the axis; $S_3$ function
    for the original pore; $S_3$ function for the rotated pore with a vector
    $(0.92, -0.4)$ being a direction of elongation of the pore; $S_3$ function
    averaged across rotations similar to the algorithm described in
    \cite{berryman1988,SMITH1988176,malmir2018}.}
  \label{fig:aniso}
\end{figure*}

@@

To summarize, \code{CorrelationFunctions.jl} allows to compute 3-point
correlation functions averaged in all directions (by randomly rotating the
sample or the pattern), along predefined number of directions (e.g., orthogonal
and diagonal by rotation in these directions) or along the direction of major
anisotropy (that can be determined using \code{$detect\_anisotropy$} function).
Full correlation map for 3-point functions is impractical due its huge size and
is not implemented.

@@

The test is performed by firstly evaluating the inequation in a cube
$[-1, 1]^3$ and obtaining precise values of $F_{sss}$ and secondly computing
$F_{sss}$ with our algorithm, giving a discretized volume of $400^3$
voxels as the input. For ease of comparison, $F_{sss}$ function is evaluated within $XY$
plane. Pointwise relative error of $F_{sss}$ is on
\cref{fig:sss-dumbbell-error}. Maximal relative error in the area distanced
farther than 0.1 (20 pixels) from curves of discontinuity (see
\cref{fig:sss-dumbbell-precise}) is no more than 20\%.
\begin{figure*}[tp]
  \centering
  \subfigure[The dumbbell-like object used in testing of $F_{sss}$ computation]{
    \includegraphics[width=0.4\linewidth]{images/dumbbell.png}
    \label{fig:sss-dumbbell}}
  \hfill
  \subfigure[Relative error of computation]{
    \includegraphics[width=0.4\linewidth]{images/dumbbell-sss-err.png}
    \label{fig:sss-dumbbell-error}}
  \vskip\baselineskip
  \subfigure[$F_{sss}$: precise. The correlation function is undefined (has type
    II discontinuity) at curves A, B and C.]{
    \includegraphics[width=0.4\linewidth]{images/dumbbell-sss-precise.png}
    \label{fig:sss-dumbbell-precise}}
  \hfill
  \subfigure[$F_{sss}$: calculated with our approach]{
    \includegraphics[width=0.4\linewidth]{images/dumbbell-sss-julia.png}}
  \caption[]{A comparison of $F_{sss}$ function calculated using our approach
    with its precise values.}
  \label{fig:sss-verification}
\end{figure*}

@@

@@

Wall times are shown in \cref{fig:timings}. It can be seen that the time of
execution is a power function of voxels number. GPU gives a
10x-100x performance boost depending on the correlation function. Note that
random pixel/voxel patterns are a hard test for some functions, especially for
$C_3$, as it produces a huge number of separate clusters. Thus, for real samples
execution times can be even smaller compared to those reported on \cref{fig:timings}.

@@

We have made execuation time comparisons of our algorithm with existing
algorithms.  However, it is also beneficial to compare some algorithmic details
against existing digital approaches -- proposed by Malmir~et~al
\cite{malmir2018} and another one by Smith and Torquato \cite{SMITH1988176}. The
first algorithm generates a triangle with fixed sides $r_1$ and $r_2$ and an
arbitrary angle $\theta$ between them and places it in the medium firstly making
a random translation and a rotation around a random vector by a random angle. By
making a large number of trials the algorithm gets a value of a three-point
correlation function at the point $(r_1, r_2, \theta)$. The second algorithm
takes a precalculated triangle $(r_1, r_2, \theta)$ from a web-like pattern and
places it in the medium without rotation, only applying a random
translation. This way the algorithm preserves orientation of triangles stored in
the pattern. The main advantage of our approach based on right triangle is in
its parallelizability and ease of implementation. A drawback is that an angle
between catheti of the sampling pattern is fixed to be $\theta = \frac{\pi}{2}$,
but it can be mitigated by choosing a random pattern or rotating the
pattern. The rotations can be random as in the Malmir et al. algorithm, or it
can be a single rotation to align catheti of the sampling triangle with the
direction of anisotropy (\cref{fig:aniso}). The latter allows computing
directional 3-point correlation functions for anisotropic structures similar to
2-point counterparts \cite{10.1063/1.4867611,EPL1}.

@@

We mentioned that classic set of correlation functions describes both the
geometry and topology of the structure at hand -- but this is to yet unknown
extent. With the advent of analytical methods \cite{CHERKASOV2024129400} to
assess information content of arbitrary descriptor for structure of any size, it
will be possible to choose the most useful set of correlation functions and
compare their information content against classical and other metrics. If
computationally effective methodology for information content assessment will be
developed, it should amend all aforementioned uncertainties and provide a clear
winner -- a plethora of different 2-point CFs or a less numerous set of
higher-order functions. Such long awaited answer would not be universal thought
and depends not only on descriptor's information content versus data volume
ratio, but also on its applicability to solve particular problems. For example,
it is well known that cluster $C_2$ function provides a lot of information in
addition to $S_2$ \cite{JiaoPNAS}, yet it's application in stochastic
reconstructions is very limited due to computational burden of cluster
relabeling during annealing \cite{PLoS_ONE}. Still, effective information
content assessment techniques will provide a possibility to account for such
additional application constraints.

>>>>

\author{Vasily~Postnicov\textsuperscript{1,3}}
\author{Marina~V.~Karsanina\textsuperscript{1,2}}
\author{Aleksey~Khlyupin\textsuperscript{1,2}}
\author{Kirill~M.~Gerke\textsuperscript{1,2}}
\email{kg@ifz.ru}

\affiliation{\textsuperscript{1}Moscow Institute of Physics and Technology,
  Dolgoprudny, 141701, Russia}
\affiliation{\textsuperscript{2}Schmidt Institute of Physics of the Earth of
  Russian Academy of Sciences, Moscow, 107031, Russia}
\affiliation{\textsuperscript{3}Dokuchaev Soil Science Institute, Moscow, 119017, Russia}

@@

\begin{abstract}
Structures, or spatial arrangements of matter and energy, including some fields
(e.g., velocity or pressure) are ubiquitous in research applications and
frequently require description for subsequent analysis, or stochastic
reconstruction from limited data.  The classical descriptors are 2-point
correlation functions (CFs), but the computation of 3-point statistics is known
to be advantageous in some cases as they can probe non-Gaussian signatures, not
captured by their 2-point counterparts. Moreover, $n$-point CFs with $n \ge 3$
are believed to possess larger information content and provide more information
about studied structures. In this paper, we have developed algorithms and code
to compute $S_3, C_3, F_{sss}, F_{ssv}$ and $F_{svv}$ with a right-angle and arbitrary
triangle pattern. The former was believed to be faster to compute, but with the help of precomputed
regular positions we achieved the same speed for arbitrary pattern. In this work we also 
implement and demonstrate computations of directional 3-point CFs -- for this
purpose right-triangular pattern seems to be superior due to explicit orientation and high coverage.
Moreover, we assess the errors in CFs' evaluation due to image or pattern rotations and show that they 
have minor effect on accuracy of computations.
The execution times of our algorithms for the same number of samples are orders of
magnitude lower than in existing published counterparts. We showed that the volume
of data produced get unwieldy very easily, especially if computations are
performed in frequency domain. For these reasons until information content of
different sets of correlation functions with different "n-pointness" is not
known, advantages of CFs with $n>3$ are not clear. Nonetheless, developed
algorithms and code are universal enough to be easily extendable to any $n$
(with increasing computational and RAM burden). All results are available as
part of open source package \code{CorrelationFunctions.jl} [Postnicov et al.,
  2024, Comp.Phys.Commun., 299, 109134]. As described in this paper, 3-point CFs
computations can be immediately applied in a great number of research
applications, for example: 1) flow and transport velocity fields analysis or any
data with non-Gaussian signatures, 2) deep learning for structural and physical
properties, 3) structure taxonomy and categorization.  In all these and numerous
other potential cases the ability to compute directional 3-point functions may
be crucial. Notably, the organization of the code functions allows computation
of cross-correlation, i.e., one can compute 3-point CFs for multiphase images
(while binary structures were used in this paper for simplicity of
explanations).
\end{abstract}

@@

\section{Introduction}
\label{sec:intro}
Structures, or spatial arrangements of matter and energy, including some fields
(e.g., velocity or pressure) are ubiquitous in research applications and
frequently require description for subsequent analysis. Examples of such objects
for studies include galaxy formations \cite{springel2006}, immiscible multiphase
fluid flow patterns \cite{hopkins2015new,balashov2021}, rock and soil samples
\cite{rozenbaum2014,karsanina2015,ledesma2018,chen2020super,prokhorov2021digital,vogel2010},
food specimens \cite{derossi2019,nagdalian2021}, biological tissue
\cite{park2020}, vortexes during flow \cite{gorbunova2016precessing} and
interstellar turbulence \cite{portillo2018developing}. The scale of interest may
span orders of magnitude from angstroms and nm for materials
\cite{garum2020,gerke2021,khlyupin2023molecular} to millions of light-years for
star clusters \cite{takada2003three,hopkins2013stars}. The studied structure can
be non-static -- it will change with time, and this dynamics is usually also of
great interest \cite{jiao2013,fomin2023soil}. Thus, description of the structure
is of utmost importance for numerous applications:
\begin{enumerate}
  \item Elucidation of structure--property relationships in the form of
    theoretical expression \cite{Torquato_book,Sahimi_book}, e.g., rigorous
    bounds and such;
  \item Prediction of material physical properties using machine (deep) learning
    \cite{obayashi2018persistence,kamrava2020linking,roding2020predicting};
  \item Description of complex structures and fields for taxonomy, comparison
    and analysis
    \cite{takada2003three,hopkins2013stars,shivashankar2015felix,saadatfar2017pore,
      portillo2018developing,cheng2022data,KarsaninaEJSS,PNM_Morse,khlyupin2023molecular},
    which may include temporal dynamics or surface evolution
    \cite{jiao2013,PhysRevE.92.023301,prokhorov2022,chen2022,fomin2023soil};
  \item Morphological representativity
    \cite{capek2009,rozenbaum2014,gerke2019tensor,zubov2024search} and spatial stationarity
    analysis \cite{REVpaper,LavrukhinPRE,zubov2024search};
  \item Stochastic reconstructions from limited amount of input data
    \cite{Adler_recon,Y-T,tahmasebiPRL,Euras2012,EPL2,karsaninaPRL,rozanski2023}
    in the form of limited dimension structures or correlation functions;
  \item Compress structural information for storage, analysis, training (feature
    extraction), and retrieval (by means of stochastic reconstructions)
    \cite{jiao2007,SciRep1,Havelka,KarsaninaEJSS};
  \item Multiscale image fusion that is usually necessary due to trade-off
    between field-of-view and imaging resolution
    \cite{SciRep1,Geoderma2018,chen2020super,yan2023multiscale,karimpouli2022,yan2024modeling}.
\end{enumerate}

The number of possible descriptors to quantify spatial structure developed
within different disciplines is stunning, for example: phase ratio or porosity
\cite{anovitz2015characterization}, surface area, radial distribution functions
\cite{zimm1948scattering,becker2010radial}, Minkowski functionals
\cite{vogel2010} and tensors \cite{schroder2011minkowski}, correlation functions
\cite{Torquato_book}, persistence diagrams
\cite{shivashankar2015felix,saadatfar2017pore,obayashi2018persistence,zubov2024search}. The
evolution of such metrics was mainly induced by the ease of measurement and
interpretation. In fact, we need a single universal descriptor defined by only
one characteristic -- information content. The latter is an ability of the
metric to describe the structure at hand with 100\% meaning the ability to
fully recover the structure back from this metric. Thus, universal descriptor
has to possess following key properties: (1) high information content, (2) ease
of assess, (3) high compression ratio (i.e., descriptor is much "smaller" than
the structure itself).

@@

is based on so-called correlation functions (CFs). Majority of these functions, with 
surface functions being a notable exception \cite{samarin2023robust}, define a

@@

    for example, to add functions into a single structure \cite{moctezuma2002};

@@

    of higher order $S_n$ are known and were found to be closer to direct

@@

\subsection{Triangular sampling pattern}
There are three major known sampling pattern implementations reported in the literature (\cref{fig:pattern}):
\begin{enumerate}
  \item Arbitrary pattern (or random) triangle is thrown into the structure (e.g., \cite{malmir2018});
  \item Web-like predefined triangular pattern (e.g., \cite{SMITH1988176});
  \item Right-triangular sampling pattern (e.g., \cite{dimitrakopoulos2010high}).
\end{enumerate}
The second case of the web-like pattern was useful at times of low computational resources and seems to be the least flexible in terms of structure coverage. Right-traingle as a pattern was popular due ease of implementation and potential speed gains due to vectorization.
However, we have organized the algorithms of \code{CorrelationFunctions.jl} to be able to utilize any patterns.
This was achieved by organizing computational functions not around a shifting pattern, but rather around precomputed points representing
triangle's vertices -- thus making our approach under the hood using the ideas behind the web-like pattern method.

This is implemented in our code by using the same interface for all computation functions to assess CFs:
\begin{verbatim}
corrfn(input, phase, ps1, ps2)
\end{verbatim}
where \code{corrfn} stands for \code{s3} ($S_3$), \code{c3} ($C_3$),
\code{surf3} ($F_{sss}$) etc., \code{input} is the input array describing the structure, \code{phase} is
a phase of interest and \code{ps1} and \code{ps2} are arrays of points in which
correlation function is calculated. Firstly, \code{corrfn} does all necessary
preparations for the input array (e.g. edge extraction or connected component
labeling) and then calculates the correlation function for each pair of
arguments from \code{ps1} and \code{ps2}.

To compare different approaches in this work we performed calculations with right-triangular\cite{dimitrakopoulos2010high,malmir2018},
equilateral triangle and isosceles triangle with varying angle \cite{SMITH1988176,malmir2018} patterns -- both with averaging
over the whole structure, or computed in directions (see next two subsections below). In the first case the pattern where arguments
of a correlation function $x_1$ and $x_2$ are fixed to be parallel to one of the
axes $X$, $Y$ or $Z$ with restriction $x_1 \perp x_2$ (\cref{fig:pattern}, right scheme). Except for equilateral triangle, the output  has dimensions $D_1 \times D_2$ where $D_1$ and $D_2$ are dimensions of the input array along selected axes. For equilateral pattern there is a single curve (i.e.,
similar to $S_2$ graphical representation).

\begin{figure*}[ht]
  \centering
  \includegraphics[width=0.8\linewidth]{images/web-like.png}
  \caption[]{Schematic representation of patterns for 3-point correlation functions evaluation:
  arbitrary or random triangle, web-like, and right-triangular patterns. Note that real patterns are
  more complex and are made simple for visualization, i.e., for right-triangular pattern we sample
  all possible variations between points. Equilateral or isosceles triangles can be considered
  as arbitrary patterns and can be either fixed in space, or rotated randomly.}
  \label{fig:pattern}
\end{figure*}

\subsection{Analysis of anisotropic media}
When an analyzed sample is anisotropic, i.e., it is stretched or oriented along
some specific direction, it may be beneficial to align one of the sides (cathetus for the right-triangular pattern) of
sampling triangle \cref{fig:pattern} with that direction. In this case the
anisotropy present in the sample will be properly captured by computed 3-point
correlation function.

To detect the direction of anisotropy we use an $N \times N$ (where $N$ is
dimensionality of the sample) covariance matrix
$\Sigma = \left\{ \sigma_{ij} \right\}$ for coordinates of points in the phase of
interest:
\begin{equation}
  \sigma_{ij} = \sum_{x \in \left\{
    \begin{array}{l}
      \text{Coordinates of points} \\
      \text{in the phase of interset}
    \end{array}
    \right\}} (x_i - \overline{x_i})(x_j - \overline{x_j})
\end{equation}
and decompose $\Sigma$ as $T \Lambda T^{-1}$ where $T$ is a unitary matrix and
$\Lambda$ is diagonal. $T$ can always be made so that it defines a rotation (by
multiplication of one of the columns by $-1$ if needed, so that
$\det T = 1$). Application of this rotation to the sample (or its inverse to the
pattern) makes the direction of anisotropy coaligned with the pattern, so
anisotropy can be better represented in the correlation function. For the sake
of visual clarity in this study we choose to rotate the image instead of the
pattern (\cref{fig:aniso}). Any rotation can potentially reduce accuracy of
computed functions due to necessity to interpolate the image during
rotation, and this important issue is explored later.
\begin{figure*}[ht]
  \centering
  \includegraphics[width=0.8\linewidth]{images/fig_rot.png}
  \caption{Examples of directional analysis of images with the help of $S_3$ function
    when dealing with highly anisotropic media. The right-triangular pattern (\cref{fig:pattern}, right side)
    which we use in our algorithm is most sensitive to anisotropy which is
    directed along one of the axes. \\
    From top down, left to right: an original image of an anisotropic pore; a
    rotated image of the pore, so it is positioned along the major axis; $S_3$ function
    for the original pore; $S_3$ function for the rotated pore with a vector
    $(0.92, -0.4)$ being a direction of elongation of the pore; $S_3$ function
    averaged across rotations similar to previously known algorithms
    \cite{berryman1988,SMITH1988176,malmir2018}.}
  \label{fig:aniso}
\end{figure*}

@@

To summarize, \code{CorrelationFunctions.jl} allows to compute 3-point
correlation functions averaged in all directions (by randomly rotating the
sample or the pattern), along predefined number of directions (e.g., orthogonal
and diagonal by rotation in these directions) or along the direction of major
anisotropy (that can be determined using \code{$detect\_anisotropy$} function).
Full correlation map for 3-point functions is impractical due its huge size and, thus,
is not implemented.

@@

The test is performed by firstly evaluating the inequation in a cube
$[-1, 1]^3$ and obtaining precise values of $F_{sss}$ and secondly computing
$F_{sss}$ with our algorithm, giving a discretized volume of $400^3$
voxels as the input. For ease of comparison, $F_{sss}$ function is evaluated within $XY$
plane. Pointwise relative error of $F_{sss}$ is on
\cref{fig:sss-dumbbell-error}. Maximal relative error in the area distanced
farther than 0.1 (20 pixels) from curves of discontinuity (see
\cref{fig:sss-dumbbell-precise}) is no more than 20\%.
\begin{figure*}[tp]
  \centering
  \subfigure[The dumbbell-like object used in testing of $F_{sss}$ computation]{
    \includegraphics[width=0.4\linewidth]{images/dumbbell.png}
    \label{fig:sss-dumbbell}}
  \hfill
  \subfigure[Relative error of computation]{
    \includegraphics[width=0.4\linewidth]{images/dumbbell-sss-err.png}
    \label{fig:sss-dumbbell-error}}
  \vskip\baselineskip
  \subfigure[$F_{sss}$: precise. The correlation function is undefined (has type
    II discontinuity) at curves A, B and C.]{
    \includegraphics[width=0.4\linewidth]{images/dumbbell-sss-precise.png}
    \label{fig:sss-dumbbell-precise}}
  \hfill
  \subfigure[$F_{sss}$: calculated with our approach]{
    \includegraphics[width=0.4\linewidth]{images/dumbbell-sss-julia.png}}
  \caption[]{A comparison of $F_{sss}$ function calculated using our algorithm
    with its precise values.}
  \label{fig:sss-verification}
\end{figure*}

@@

\subsection{Comparison of sampling patterns}
\label{sec:compare_patterns}
To compare different sampling and averaging patterns we performed computations of $S_3$ on Sandstone 3D sample.
They were performed in three different manners: 1) isosceles triangle with a right-triangle
between the equal side -- this sampling resembles both our right-triangle pattern and pattern
of Malmir et al. \cite{malmir2018}, computations were made within XY planes within the image and, thus,
represent directional CF; 2) equilateral triangle with side $r$ directionally analyzed within XY planes;
3) equilateral triangle with side $r$ averaged over 20 random rotations -- reminiscence of web-like and random patterns.

The results of the comparison are shown in \cref{fig:comparison_patterns}. For isotropic samples curves for
directional and ensemble averaged, as well as for right-angle isosceles triangle, basically collapse to each other
with minimal differences.
\begin{figure*}[ht]
  \centering
  \includegraphics[width=0.8\linewidth]{images/fig_compar.png}
  \caption[]{A comparison of computations of $S_3$ correlation function for the Sandstone sample
  using different patterns with and without random rotations. Left: evaluated with isosceles triangle 
  pattern with the side $r$ and the angle between edges $\phi$; sampling triangles are aligned with the plane XY.
  Middle: evaluated with equilateral triangle pattern with the side $r$; sampling triangles are aligned with the plane XY.
  Right: evaluated with equilateral triangle pattern with the side $r$. The result is averaged over 20 
  random rotations of the sampling triangle. Lower image: comparison of $S_3$ functions calculated with 
  right triangle pattern with length of the both catheti $r$ and equilateral triangle pattern 
  (both averaged over rotations and aligned with the plane XY).}
  \label{fig:comparison_patterns}
\end{figure*}

A Jupiter notebook presenting all these cases is available online \cite{Repo_example} and is a good example of
applications of our package, especially in terms of establishing \code{ps1} and \code{ps2}
arrays for arbitrary triangle patterns (isosceles and equilateral cases in our example),
and for directional and ensemble averaged calculations.

@@

Wall times are shown in \cref{fig:timings}. They do not depend on the type of sampling pattern used. It can be seen that the time of
execution is a power function of voxels number. GPU gives a
10x-100x performance boost depending on the correlation function. Note that
random pixel/voxel patterns are a hard test for some functions, especially for
$C_3$, as it produces a huge number of separate clusters. Thus, for real samples
execution times can be even smaller compared to those reported on \cref{fig:timings}.

@@

We have made execution time comparisons of our algorithm with existing
algorithms.  However, it is also beneficial to compare some algorithmic details
against existing digital approaches -- proposed by Malmir~et.~al.
\cite{malmir2018} and another one by Smith and Torquato \cite{SMITH1988176}. The
first algorithm generates a triangle with fixed sides $r_1$ and $r_2$ and an
arbitrary angle $\theta$ between them and places it in the medium firstly making
a random translation and a rotation around a random vector by a random angle. By
making a large number of trials the algorithm gets a value of a three-point
correlation function at the point $(r_1, r_2, \theta)$. The second algorithm
takes a precalculated triangle $(r_1, r_2, \theta)$ from a web-like pattern and
places it in the medium without rotation, only applying a random
translation. This way the algorithm preserves orientation of triangles stored in
the pattern. The main advantage of our approach that can operate on arbitrary patterns is in
its parallelizability and ease of implementation. The rotations necessary
for arbitrary placement of the pattern within the image can be random as in the Malmir et al. algorithm, or it
can be a single rotation to align the sampling triangle with the
direction of anisotropy (\cref{fig:aniso}). The latter allows computing
directional 3-point correlation functions for anisotropic structures similar to
2-point counterparts \cite{10.1063/1.4867611,EPL1}.

@@

The question if image rotation affects the quality of computations remains open.
To answer this question we performed a study on value noise images.
In \cref{sec:rotqua} we demonstrate that even nearest-neighbor
interpolation results only in insignificant loss of quality. Rotation of the
pattern has same drawbacks related to interpolation of the sample which are most
noticeable at small correlation lengths. The both methods (rotation of the
sample or the pattern) produce nearly equivalent result if the sample is
homogeneous and an average pore size is much less than the sample's size (see
\cref{sec:rotrot}).

Another interesting question is which sampling pattern is more powerful?
Our results shown in \cref{fig:comparison_patterns} clear illustrate that
for isotropic structure all sampling techniques provide quite similar results.
The differences, of course, will be significant for isotropic cases and
directional 3-point CFs calculations. Again, the clear winner here is not obvious.
Right-triangular pattern allows to systematically sample the structure with
predefined direction. On the other hand, equilateral triangle allows to
minimize the volume of computed functions and represent them as single curve,
similar to 2-point functions. Random pattern could be useful for rigorous bounds.
So, the better approach can depend on the problem at hand. Nonetheless,
the direct comparison of patterns can be performed only on the basis of
information content -- something we discuss in the next paragraph.

We mentioned that classic set of correlation functions describes both the
geometry and topology of the structure at hand -- but this is to yet unknown
extent. With the advent of analytical methods \cite{CHERKASOV2024129400} to
assess information content of arbitrary descriptor for structure of any size, it
will be possible to choose the most useful set of correlation functions and
compare their information content against classical and other metrics. If
computationally effective methodology for information content assessment will be
developed, it should amend all aforementioned uncertainties and provide a clear
winner -- a plethora of different 2-point CFs or a less numerous set of
higher-order functions. Such long awaited answer would not be universal thought
and depends not only on descriptor's information content versus data volume
ratio, but also on its applicability to solve particular problems. For example,
it is well known that cluster $C_2$ function provides a lot of information in
addition to $S_2$ \cite{JiaoPNAS}, yet it's application in stochastic
reconstructions is very limited due to computational burden of cluster
relabeling during annealing \cite{karsanina2015}. Still, effective information
content assessment techniques will provide a possibility to account for such
additional application constraints.
